\chapter{EXPERIMENTAL STUDY}

This chapter presents the complete set of experimental results produced by the platform described in Chapter~3. The results cover 92 unique architecture--model configurations evaluated across five publicly available benchmarks, totalling 20{,}000 scored samples; runs producing incomplete \texttt{Response} objects were excluded, as were exploratory blackboard-family runs executed off the principal operating point (Section~\ref{sec:benchmark-results}). Every reported configuration corresponds to a single benchmark run at its principal operating point, and together they cover the full architecture taxonomy defined in Chapter~3: the three architectures that were pending in the previous evaluation snapshot---Debate (POA), Blackboard, and Entropy Blackboard---have since been benchmarked on all five benchmarks under the same evaluation protocol. Section~\ref{sec:eval-env} summarizes the evaluation environment. Section~\ref{sec:bench-suite} defines the benchmark suite and evaluation protocol. Section~\ref{sec:benchmark-results} presents per-benchmark result tables for every evaluated configuration. Section~\ref{sec:arch-analysis} provides an architecture-family analysis synthesizing findings across benchmarks. Section~\ref{sec:cross-bench} reports a cross-benchmark EATS comparison. Section~\ref{sec:exp-discussion} discusses the principal findings and noteworthy observations. Section~\ref{sec:pending} describes planned extensions.

\section{Evaluation Environment}
\label{sec:eval-env}

The platform uses a split deployment model in which the control plane runs locally while model inference runs on remote GPU hosts. Table~\ref{tab:environment-matrix} summarizes the environment.

\begin{table}[htbp]
    \centering
    \caption{Experimental environment and its role in the platform.}
    \label{tab:environment-matrix}
    \small
    \renewcommand{\arraystretch}{1.15}
    \begin{tabularx}{\linewidth}{@{} >{\raggedright\arraybackslash}p{0.21\linewidth} >{\raggedright\arraybackslash}p{0.25\linewidth} >{\raggedright\arraybackslash}X @{}}
        \toprule
        \textbf{Layer} & \textbf{Technology} & \textbf{Role in the experiment stack} \\
        \midrule
        Local control plane & FastAPI, Next.js, MLflow, Python runner & Experiment launch, tracking, result storage, and UI management. \\
        Dedicated SLM serving & Dockerized vLLM on Google Cloud L4 machines & Low-cost SLM inference for baseline, hybrid, and swarm components. \\
        Shared mid-tier serving & Dockerized vLLM on a Google Cloud RTX PRO6000 host & Shared 20B--40B serving for escalation and stronger baselines. \\
        Heavy serving & Dockerized vLLM on a Nebius H200 host & Frontier-scale serving for the largest baselines and fallback paths. \\
        Measurement layer & Metrics proxy, resource profiles, MLflow, JSON/Markdown reporter & Latency, token, cost, energy, carbon, and experiment summary collection. \\
        \bottomrule
    \end{tabularx}
\end{table}

All model endpoints are exposed through OpenAI-compatible HTTP interfaces so that the experiment runner remains provider-neutral. The platform contains host-lock logic to prevent incompatible shared-host models from running concurrently, which would otherwise contaminate latency and energy measurements.

\section{Benchmark Suite and Evaluation Protocol}
\label{sec:bench-suite}

Five benchmarks are evaluated in this chapter. All runs use zero temperature. Accuracy is the proportion of samples for which the predicted answer matches the reference. Individual experiment artefacts use the per-experiment normalisation mode described in Section~\ref{sec:normalisation}, which normalises each run against the registry-based full-LLM baseline for the exact same benchmark, sample count, and seed. For the aggregated benchmark tables in this chapter, EATS is recomputed in aggregate mode---using the mean of all standalone monolithic runs on that benchmark as the common reference, as defined in Section~\ref{sec:normalisation}---so that heterogeneous standalone, hybrid, ensemble, and swarm configurations are placed on the same per-benchmark comparison scale. For GSM8K, an early version of the numeric answer-extraction heuristic failed to parse the output format of some model families; the extraction logic was subsequently corrected and all affected GSM8K runs were rescored from their stored raw responses, so the GSM8K accuracies reported here reflect the corrected scoring (Section~\ref{sec:exp-discussion}).

\begin{table}[htbp]
    \centering
    \caption{Benchmark suite used in the experimental evaluation.}
    \label{tab:benchmark-suite}
    \small
    \begin{tabularx}{\linewidth}{@{} l l l X @{}}
        \toprule
        \textbf{Benchmark} & \textbf{Domain} & \textbf{Task type} & \textbf{Scoring and coverage} \\
        \midrule
        MMLU & World knowledge & Multiple-choice (4-way) & Exact-match accuracy; 57 academic subjects across STEM, humanities, and social sciences. \\
        ARC-Challenge & Science reasoning & Multiple-choice (4-way) & Exact-match accuracy on the ARC Challenge split; multi-step physical-world inference. \\
        GSM8K & Mathematical reasoning & Free-form (numeric) & Exact-match accuracy on the numeric answer; grade-school arithmetic and word problems. \\
        HellaSwag & Commonsense reasoning & Multiple-choice (4-way) & Exact-match accuracy; context-based sentence completion requiring real-world inference. \\
        TruthfulQA & Factual calibration & Multiple-choice & Exact-match accuracy; selects factually accurate answers over plausible-sounding falsehoods. \\
        \bottomrule
    \end{tabularx}
\end{table}

\section{Benchmark-Level Results}
\label{sec:benchmark-results}

The following tables report aggregated results for every evaluated architecture--model combination that meets the reporting threshold of at least 100 total evaluated samples. Rows are sorted alphabetically by architecture, then by model key. The \emph{N} column is the total number of samples evaluated across all runs of that configuration. Column \emph{LLM\,\%} is the fraction of queries that invoked the heavy language model. Single-model ensemble configurations (e.g., \texttt{ensemble\,|\,ministral3-3b}) represent a single SLM processed through the ensemble contract with no voting; they are included as lean efficiency references rather than true multi-model aggregations. Debate (POA) configurations are reported separately by arbitrator type: \emph{SLM arb.} denotes an all-SLM debate, while \emph{LLM arb.} denotes a debate whose final arbitration is delegated to the LLM. The label \texttt{entropy\_bb} abbreviates Entropy Blackboard. Blackboard and Entropy Blackboard rows report the SLM-worker\,$\rightarrow$\,LLM-sweeper pair, and their \emph{LLM\,\%} column is the fraction of queries resolved by the LLM sweeper. These rows are further separated by claim policy: under \emph{first threshold}, the first worker whose bid clears the bid threshold claims the task, whereas under \emph{highest bid}, the task is awarded by auction to the top eligible bidder once all workers have posted bids. To keep blackboard-family comparisons fair, every reported Blackboard and Entropy Blackboard row is a single 100-sample run at the principal operating point (bid threshold 0.75, 3.5\,s time-to-live, entropy weight 0.5 where applicable). Exploratory runs off this operating point---a bid-threshold exploration on MMLU and legacy runs executed before the claim policy became configurable, at bid threshold 0.65 and a 1.5\,s time-to-live---are excluded from the tables and summarised in Section~\ref{sec:arch-analysis}. The design--policy matrix of \{Blackboard, Entropy Blackboard\} $\times$ \{first threshold, highest bid\} is now complete on all five benchmarks at the principal operating point.

\subsection{MMLU}

Twenty-one configurations meet the reporting threshold on MMLU, covering all nine architecture families. The best standalone checkpoint \texttt{gemma4-31b} attains the highest reported accuracy (87.8\%), with \texttt{qwen3.5-27b} (87.2\%) and \texttt{gpt-oss-120b} (87.0\%) close behind. The efficiency picture is led by the hybrid pair: speculative decoding reaches 82.0\% accuracy at the highest EATS of any configuration on MMLU (0.823), and routing follows at 77.0\% accuracy (EATS 0.798) while limiting heavy-model usage to 56\% of queries. Among standalone checkpoints the MoE model \texttt{qwen3.5-35b-a3b} achieves the highest standalone EATS (0.782), ahead of \texttt{gemma4-26b-a4b} (0.759) and \texttt{gpt-oss-20b} (0.744). The decentralised architectures land in the middle of the field: Entropy Blackboard reaches 82.0--83.0\% accuracy under its two claim policies (EATS 0.685--0.692), three to six points above the standard Blackboard variants (77.0--79.0\%, EATS 0.660--0.685) at similar escalation rates. LLM-arbitrated Debate reaches 80.0\% accuracy while its all-SLM variant falls to 69.0\%, and Pure Swarm reaches 75.0\% using no heavy-model calls.

\begin{table}[htbp]
\centering
\caption{MMLU results across 21 configurations meeting the reporting threshold. Routing and speculative latency values are algorithmic (model-side inference only).}
\label{tab:result-mmlu}
\scriptsize
\renewcommand{\arraystretch}{1.15}
\begin{tabularx}{\linewidth}{@{} l X r r r r r r r @{}}
\toprule
\textbf{Arch.} & \textbf{Model} & \textbf{N} & \textbf{Acc\,\%} & \textbf{Lat\,(s)} & \textbf{\$/q} & \textbf{g\,CO$_2$/q} & \textbf{LLM\,\%} & \textbf{EATS} \\
\midrule
active\_oracle & gemma4-4b $\rightarrow$ gemma4-31b & 100 & 72.0 & 10.2 & 0.00221 & 0.083 & 53 & 0.670 \\
blackboard & gemma4-4b $\rightarrow$ gemma4-31b (first threshold) & 100 & 79.0 & 11.4 & 0.00312 & 0.131 & 31 & 0.685 \\
blackboard & gemma4-4b $\rightarrow$ gemma4-31b (highest bid) & 100 & 77.0 & 12.4 & 0.00344 & 0.145 & 31 & 0.660 \\
debate & gemma4-4b (SLM arb.) & 100 & 69.0 & 25.6 & 0.00532 & 0.194 & 0 & 0.517 \\
debate & gemma4-4b $\rightarrow$ gemma4-31b (LLM arb.) & 100 & 80.0 & 17.9 & 0.00714 & 0.337 & 100 & 0.576 \\
ensemble & gemma4-4b + llama3.2-3b + ministral3-3b & 100 & 59.0 & 26.9 & 0.00561 & 0.205 & 0 & 0.447 \\
ensemble & gemma4-4b + llama3.2-3b + ministral3-3b + qwen3.5-4b & 100 & 68.0 & 21.6 & 0.00942 & 0.344 & 0 & 0.469 \\
ensemble & ministral3-3b & 100 & 57.0 & 3.5 & 0.00074 & 0.027 & 0 & 0.639 \\
ensemble & qwen3.5-4b & 100 & 75.0 & 17.6 & 0.00367 & 0.134 & 0 & 0.616 \\
entropy\_bb & gemma4-4b $\rightarrow$ gemma4-31b (first threshold) & 100 & 82.0 & 12.3 & 0.00340 & 0.143 & 33 & 0.692 \\
entropy\_bb & gemma4-4b $\rightarrow$ gemma4-31b (highest bid) & 100 & 83.0 & 13.4 & 0.00369 & 0.155 & 37 & 0.685 \\
pure\_swarm & gemma4-4b & 100 & 75.0 & 12.3 & 0.00256 & 0.094 & 0 & 0.669 \\
routing & gemma4-4b $\rightarrow$ gemma4-31b & 100 & 77.0 & 2.5 & 0.00081 & 0.036 & 56 & 0.798 \\
speculative & gemma4-4b $\rightarrow$ gemma4-31b & 100 & 82.0 & 2.5 & 0.00107 & 0.052 & 100 & 0.823 \\
standalone & gemma4-26b-a4b & 500 & 86.4 & 5.2 & 0.00341 & 0.176 & 100 & 0.759 \\
standalone & gemma4-31b & 500 & 87.8 & 9.6 & 0.00626 & 0.324 & 100 & 0.673 \\
standalone & gpt-oss-120b & 500 & 87.0 & 3.4 & 0.00838 & 0.214 & 100 & 0.675 \\
standalone & gpt-oss-20b & 500 & 79.8 & 3.5 & 0.00299 & 0.120 & 100 & 0.744 \\
standalone & llama3.3-70b & 500 & 84.8 & 11.1 & 0.02034 & 0.700 & 100 & 0.464 \\
standalone & qwen3.5-27b & 500 & 87.2 & 19.3 & 0.01258 & 0.651 & 100 & 0.525 \\
standalone & qwen3.5-35b-a3b & 500 & 86.2 & 4.3 & 0.00279 & 0.144 & 100 & 0.782 \\
\bottomrule
\end{tabularx}
\end{table}

\subsection{ARC}

Eighteen configurations were evaluated on ARC. The standalone \texttt{gemma4-31b} attains the highest accuracy (97.0\%), ahead of a tight cluster of 95.0--95.4\% checkpoints. The efficiency frontier, however, is dominated by the hybrid pair: speculative decoding reaches 94.0\% accuracy at EATS 0.908---the highest EATS recorded across any configuration on any benchmark in this study---and routing reaches 91.0\% at EATS 0.904, at algorithmic latencies of 1.4--1.7 seconds and per-query costs of \$0.0005--\$0.0008. Active Oracle is a close third on this benchmark (91.0\% accuracy, EATS 0.886): its targeted sub-query mechanism resolves most items on the SLM at only 2.8 seconds of algorithmic latency. These three hybrids all improve on the best standalone option (\texttt{qwen3.5-35b-a3b}, EATS 0.859) while operating at 2--5$\times$ lower cost. ARC is also strong for the blackboard family: the highest-bid Blackboard reaches 95.0\% accuracy with only 13\% sweeper escalation (EATS 0.808), its first-threshold counterpart reaches 92.0\% (EATS 0.800), and Entropy Blackboard reaches 92.0--93.0\% under its two claim policies (EATS 0.791--0.799). Pure Swarm reaches 92.0\% with no heavy-model calls (EATS 0.818), and LLM-arbitrated Debate matches the strongest 31B-class standalone accuracy at 95.0\%.

\begin{table}[htbp]
\centering
\caption{ARC results across 18 configurations. Routing and speculative latency values are algorithmic (model-side inference only).}
\label{tab:result-arc}
\scriptsize
\renewcommand{\arraystretch}{1.15}
\begin{tabularx}{\linewidth}{@{} l X r r r r r r r @{}}
\toprule
\textbf{Arch.} & \textbf{Model} & \textbf{N} & \textbf{Acc\,\%} & \textbf{Lat\,(s)} & \textbf{\$/q} & \textbf{g\,CO$_2$/q} & \textbf{LLM\,\%} & \textbf{EATS} \\
\midrule
active\_oracle & gemma4-4b $\rightarrow$ gemma4-31b & 100 & 91.0 & 2.8 & 0.00060 & 0.022 & 6 & 0.886 \\
blackboard & gemma4-4b $\rightarrow$ gemma4-31b (first threshold) & 100 & 92.0 & 7.7 & 0.00180 & 0.070 & 11 & 0.800 \\
blackboard & gemma4-4b $\rightarrow$ gemma4-31b (highest bid) & 100 & 95.0 & 8.3 & 0.00194 & 0.075 & 13 & 0.808 \\
debate & gemma4-4b (SLM arb.) & 100 & 90.0 & 18.3 & 0.00382 & 0.139 & 0 & 0.659 \\
debate & gemma4-4b $\rightarrow$ gemma4-31b (LLM arb.) & 100 & 95.0 & 12.7 & 0.00521 & 0.248 & 100 & 0.679 \\
ensemble & gemma4-4b + llama3.2-3b + ministral3-3b + qwen3.5-4b & 100 & 91.0 & 15.9 & 0.00698 & 0.254 & 0 & 0.610 \\
entropy\_bb & gemma4-4b $\rightarrow$ gemma4-31b (first threshold) & 100 & 92.0 & 8.0 & 0.00200 & 0.080 & 15 & 0.791 \\
entropy\_bb & gemma4-4b $\rightarrow$ gemma4-31b (highest bid) & 100 & 93.0 & 7.9 & 0.00196 & 0.078 & 15 & 0.799 \\
pure\_swarm & gemma4-4b & 100 & 92.0 & 7.0 & 0.00147 & 0.054 & 0 & 0.818 \\
routing & gemma4-4b $\rightarrow$ gemma4-31b & 100 & 91.0 & 1.4 & 0.00052 & 0.024 & 75 & 0.904 \\
speculative & gemma4-4b $\rightarrow$ gemma4-31b & 100 & 94.0 & 1.7 & 0.00075 & 0.036 & 100 & 0.908 \\
standalone & gemma4-26b-a4b & 500 & 95.4 & 7.4 & 0.00482 & 0.250 & 100 & 0.723 \\
standalone & gemma4-31b & 500 & 97.0 & 7.8 & 0.00507 & 0.262 & 100 & 0.722 \\
standalone & gpt-oss-120b & 500 & 95.4 & 2.7 & 0.00620 & 0.170 & 100 & 0.734 \\
standalone & gpt-oss-20b & 500 & 93.8 & 2.7 & 0.00210 & 0.091 & 100 & 0.843 \\
standalone & llama3.3-70b & 100 & 92.0 & 8.9 & 0.01630 & 0.562 & 100 & 0.496 \\
standalone & qwen3.5-27b & 100 & 95.0 & 13.9 & 0.00905 & 0.468 & 100 & 0.588 \\
standalone & qwen3.5-35b-a3b & 500 & 95.4 & 2.8 & 0.00185 & 0.096 & 100 & 0.859 \\
\bottomrule
\end{tabularx}
\end{table}

\subsection{GSM8K}

Eighteen configurations were evaluated on GSM8K. After the answer-extraction correction and rescoring described in Section~\ref{sec:bench-suite}, the standalone models cluster tightly between 85.0\% and 96.4\%, with \texttt{qwen3.5-35b-a3b} (96.4\%) and \texttt{gpt-oss-120b} (96.2\%) leading. Pure Swarm achieves the highest EATS on this benchmark (0.841 at 94.0\% accuracy): running its coordination path entirely on \texttt{gemma4-4b} at 7.1 seconds of algorithmic latency, it is the most efficient configuration on GSM8K. The highest-bid Blackboard and the first-threshold Entropy Blackboard both reach 97.0\%---the highest raw accuracy recorded in the study, matched only by the standalone \texttt{gemma4-31b} on ARC. The first-threshold Entropy Blackboard is also among the most efficient coordination designs here, yielding EATS 0.826 with only 7\% sweeper escalation---essentially tied with the standalone \texttt{qwen3.5-35b-a3b} (0.824) and ahead of the first-threshold standard Blackboard (96.0\%, EATS 0.818 at 8\% escalation). GSM8K routing escalates only 1\% of queries to the LLM and reaches 95.0\% accuracy (EATS 0.776), indicating that \texttt{gemma4-4b} is self-sufficient on nearly all GSM8K questions at this confidence threshold; the four-SLM ensemble and LLM-arbitrated Debate also reach 96.0\%.

\begin{table}[htbp]
\centering
\caption{GSM8K results across 18 configurations. Routing and speculative latency values are algorithmic.}
\label{tab:result-gsm8k}
\scriptsize
\renewcommand{\arraystretch}{1.15}
\begin{tabularx}{\linewidth}{@{} l X r r r r r r r @{}}
\toprule
\textbf{Arch.} & \textbf{Model} & \textbf{N} & \textbf{Acc\,\%} & \textbf{Lat\,(s)} & \textbf{\$/q} & \textbf{g\,CO$_2$/q} & \textbf{LLM\,\%} & \textbf{EATS} \\
\midrule
active\_oracle & gemma4-4b $\rightarrow$ gemma4-31b & 100 & 92.0 & 12.9 & 0.00269 & 0.098 & 10 & 0.750 \\
blackboard & gemma4-4b $\rightarrow$ gemma4-31b (first threshold) & 100 & 96.0 & 9.1 & 0.00205 & 0.078 & 8 & 0.818 \\
blackboard & gemma4-4b $\rightarrow$ gemma4-31b (highest bid) & 100 & 97.0 & 11.8 & 0.00268 & 0.103 & 17 & 0.785 \\
debate & gemma4-4b (SLM arb.) & 100 & 95.0 & 24.7 & 0.00514 & 0.187 & 0 & 0.645 \\
debate & gemma4-4b $\rightarrow$ gemma4-31b (LLM arb.) & 100 & 96.0 & 21.5 & 0.00701 & 0.312 & 100 & 0.626 \\
ensemble & gemma4-4b + llama3.2-3b + ministral3-3b + qwen3.5-4b & 100 & 96.0 & 16.1 & 0.00912 & 0.333 & 0 & 0.619 \\
entropy\_bb & gemma4-4b $\rightarrow$ gemma4-31b (first threshold) & 100 & 97.0 & 9.2 & 0.00196 & 0.072 & 7 & 0.826 \\
entropy\_bb & gemma4-4b $\rightarrow$ gemma4-31b (highest bid) & 100 & 95.0 & 12.1 & 0.00269 & 0.102 & 28 & 0.772 \\
pure\_swarm & gemma4-4b & 100 & 94.0 & 7.1 & 0.00148 & 0.054 & 0 & 0.841 \\
routing & gemma4-4b $\rightarrow$ gemma4-31b & 100 & 95.0 & 12.1 & 0.00255 & 0.094 & 1 & 0.776 \\
speculative & gemma4-4b $\rightarrow$ gemma4-31b & 100 & 95.0 & 9.4 & 0.00500 & 0.251 & 100 & 0.725 \\
standalone & gemma4-26b-a4b & 500 & 94.8 & 4.6 & 0.00302 & 0.156 & 100 & 0.811 \\
standalone & gemma4-31b & 500 & 95.4 & 8.9 & 0.00582 & 0.301 & 100 & 0.709 \\
standalone & gpt-oss-120b & 500 & 96.2 & 4.1 & 0.01052 & 0.261 & 100 & 0.663 \\
standalone & gpt-oss-20b & 500 & 94.8 & 3.9 & 0.00336 & 0.133 & 100 & 0.812 \\
standalone & llama3.3-70b & 100 & 85.0 & 9.1 & 0.01673 & 0.577 & 100 & 0.487 \\
standalone & qwen3.5-27b & 100 & 95.0 & 14.8 & 0.00963 & 0.498 & 100 & 0.600 \\
standalone & qwen3.5-35b-a3b & 500 & 96.4 & 4.5 & 0.00294 & 0.152 & 100 & 0.824 \\
\bottomrule
\end{tabularx}
\end{table}

\subsection{HellaSwag}

Eighteen configurations were evaluated on HellaSwag. Speculative decoding achieves 92.0\% accuracy at EATS 0.899 and routing 90.0\% at EATS 0.893; both use algorithmic latency under 2 seconds and per-query costs below \$0.001. These results match or approach the standalone accuracy range of 78.0--92.6\% while reducing cost and latency by factors of 3--15 relative to the most expensive standalone options; the best standalone EATS is \texttt{qwen3.5-35b-a3b} (0.807), which is also the accuracy leader (92.6\%). Active Oracle reaches only 67.0\% accuracy, the lowest of any configuration on this benchmark; its EATS (0.674) is now competitive with the standalone field after its algorithmic latency was reduced to 5.8 seconds, but the residual accuracy gap reflects that its oracle sub-query mechanism does not recover the answer chain for commonsense completion tasks. HellaSwag is the benchmark where the entropy-aware bidding rule pays off most clearly: Entropy Blackboard escalates 82\% of queries to the sweeper and reaches 84.0\% accuracy (EATS 0.695), five to seven points above the two standard Blackboard variants (79.0\% and 77.0\%, both at 41\% escalation). Its two claim policies produced identical claim and escalation decisions on every HellaSwag query at zero temperature, so the two rows coincide. LLM-arbitrated Debate reaches 90.0\%, matching routing accuracy, while its all-SLM variant achieves only 72.0\% and Pure Swarm trails at 65.0\%.

\begin{table}[htbp]
\centering
\caption{HellaSwag results across 18 configurations. Routing and speculative latency values are algorithmic.}
\label{tab:result-hellaswag}
\scriptsize
\renewcommand{\arraystretch}{1.15}
\begin{tabularx}{\linewidth}{@{} l X r r r r r r r @{}}
\toprule
\textbf{Arch.} & \textbf{Model} & \textbf{N} & \textbf{Acc\,\%} & \textbf{Lat\,(s)} & \textbf{\$/q} & \textbf{g\,CO$_2$/q} & \textbf{LLM\,\%} & \textbf{EATS} \\
\midrule
active\_oracle & gemma4-4b $\rightarrow$ gemma4-31b & 100 & 67.0 & 5.8 & 0.00124 & 0.046 & 20 & 0.674 \\
blackboard & gemma4-4b $\rightarrow$ gemma4-31b (first threshold) & 100 & 77.0 & 8.5 & 0.00286 & 0.129 & 41 & 0.672 \\
blackboard & gemma4-4b $\rightarrow$ gemma4-31b (highest bid) & 100 & 79.0 & 8.6 & 0.00286 & 0.128 & 41 & 0.684 \\
debate & gemma4-4b (SLM arb.) & 100 & 72.0 & 21.7 & 0.00452 & 0.165 & 0 & 0.532 \\
debate & gemma4-4b $\rightarrow$ gemma4-31b (LLM arb.) & 100 & 90.0 & 13.2 & 0.00580 & 0.280 & 100 & 0.643 \\
ensemble & gemma4-4b + llama3.2-3b + ministral3-3b + qwen3.5-4b & 100 & 68.0 & 17.6 & 0.00691 & 0.252 & 0 & 0.486 \\
entropy\_bb & gemma4-4b $\rightarrow$ gemma4-31b (first threshold) & 100 & 84.0 & 7.6 & 0.00370 & 0.182 & 82 & 0.695 \\
entropy\_bb & gemma4-4b $\rightarrow$ gemma4-31b (highest bid) & 100 & 84.0 & 7.6 & 0.00370 & 0.182 & 82 & 0.695 \\
pure\_swarm & gemma4-4b & 100 & 65.0 & 10.1 & 0.00210 & 0.077 & 0 & 0.604 \\
routing & gemma4-4b $\rightarrow$ gemma4-31b & 100 & 90.0 & 1.6 & 0.00061 & 0.028 & 91 & 0.893 \\
speculative & gemma4-4b $\rightarrow$ gemma4-31b & 100 & 92.0 & 1.7 & 0.00072 & 0.034 & 100 & 0.899 \\
standalone & gemma4-26b-a4b & 500 & 83.2 & 7.3 & 0.00479 & 0.248 & 100 & 0.661 \\
standalone & gemma4-31b & 500 & 88.8 & 6.9 & 0.00448 & 0.232 & 100 & 0.705 \\
standalone & gpt-oss-120b & 500 & 80.6 & 3.6 & 0.00901 & 0.225 & 100 & 0.593 \\
standalone & gpt-oss-20b & 500 & 78.0 & 3.8 & 0.00325 & 0.128 & 100 & 0.703 \\
standalone & llama3.3-70b & 100 & 89.0 & 7.4 & 0.01380 & 0.472 & 100 & 0.531 \\
standalone & qwen3.5-27b & 100 & 92.0 & 14.4 & 0.00942 & 0.487 & 100 & 0.573 \\
standalone & qwen3.5-35b-a3b & 500 & 92.6 & 4.0 & 0.00261 & 0.135 & 100 & 0.807 \\
\bottomrule
\end{tabularx}
\end{table}

\subsection{TruthfulQA}

Seventeen configurations were evaluated on TruthfulQA. Speculative decoding attains both the highest accuracy (88.0\%) and the highest EATS (0.860) on this benchmark, with routing close behind (85.0\% accuracy, EATS 0.854); both operate at sub-2-second algorithmic latency. The strongest standalone checkpoint is \texttt{gemma4-31b} (87.2\% accuracy, EATS 0.704), while the most efficient standalone is \texttt{qwen3.5-35b-a3b} (EATS 0.716). The four-SLM ensemble achieves only 64.0\%, substantially lower than on other benchmarks, suggesting that majority-vote coordination among small models is ineffective when individual models' misconceptions are correlated. TruthfulQA is the weakest benchmark for the blackboard family at the principal operating point: standard Blackboard lands at 68.0--69.0\% under both claim policies, while Entropy Blackboard recovers part of that loss (73.0--75.0\% at 83--84\% escalation). The two standard-Blackboard policies are within one point of each other here, confirming that the claim rule is not what limits performance; an exploratory legacy first-threshold run, excluded from the table, reached 83.0\% only by escalating every query to the sweeper---effectively LLM fallback driven by its short time-to-live (Section~\ref{sec:arch-analysis}). All blackboard-family configurations remain below the routing configuration built from the same model pair (85.0\%). LLM-arbitrated Debate performs notably better at 87.0\%, the third-highest accuracy on this benchmark, suggesting that adversarial critique is more effective than task-board escalation at filtering plausible-sounding falsehoods.

\begin{table}[htbp]
\centering
\caption{TruthfulQA results across 17 configurations. Routing and speculative latency values are algorithmic.}
\label{tab:result-truthfulqa}
\scriptsize
\renewcommand{\arraystretch}{1.15}
\begin{tabularx}{\linewidth}{@{} l X r r r r r r r @{}}
\toprule
\textbf{Arch.} & \textbf{Model} & \textbf{N} & \textbf{Acc\,\%} & \textbf{Lat\,(s)} & \textbf{\$/q} & \textbf{g\,CO$_2$/q} & \textbf{LLM\,\%} & \textbf{EATS} \\
\midrule
active\_oracle & gemma4-4b $\rightarrow$ gemma4-31b & 100 & 72.0 & 8.6 & 0.00198 & 0.077 & 89 & 0.636 \\
blackboard & gemma4-4b $\rightarrow$ gemma4-31b (first threshold) & 100 & 68.0 & 11.8 & 0.00333 & 0.141 & 44 & 0.547 \\
blackboard & gemma4-4b $\rightarrow$ gemma4-31b (highest bid) & 100 & 69.0 & 13.0 & 0.00366 & 0.155 & 46 & 0.537 \\
debate & gemma4-4b (SLM arb.) & 100 & 72.0 & 18.7 & 0.00389 & 0.142 & 0 & 0.517 \\
debate & gemma4-4b $\rightarrow$ gemma4-31b (LLM arb.) & 100 & 87.0 & 12.0 & 0.00504 & 0.241 & 100 & 0.610 \\
ensemble & gemma4-4b + llama3.2-3b + ministral3-3b + qwen3.5-4b & 100 & 64.0 & 15.6 & 0.00639 & 0.233 & 0 & 0.442 \\
entropy\_bb & gemma4-4b $\rightarrow$ gemma4-31b (first threshold) & 100 & 75.0 & 9.6 & 0.00346 & 0.159 & 83 & 0.601 \\
entropy\_bb & gemma4-4b $\rightarrow$ gemma4-31b (highest bid) & 100 & 73.0 & 7.8 & 0.00310 & 0.146 & 84 & 0.612 \\
pure\_swarm & gemma4-4b & 100 & 67.0 & 8.8 & 0.00184 & 0.067 & 0 & 0.606 \\
routing & gemma4-4b $\rightarrow$ gemma4-31b & 100 & 85.0 & 1.5 & 0.00057 & 0.027 & 81 & 0.854 \\
speculative & gemma4-4b $\rightarrow$ gemma4-31b & 100 & 88.0 & 1.7 & 0.00076 & 0.036 & 100 & 0.860 \\
standalone & gemma4-26b-a4b & 500 & 84.8 & 6.7 & 0.00436 & 0.225 & 100 & 0.653 \\
standalone & gemma4-31b & 500 & 87.2 & 5.4 & 0.00350 & 0.181 & 100 & 0.704 \\
standalone & gpt-oss-120b & 500 & 82.4 & 3.0 & 0.00710 & 0.189 & 100 & 0.613 \\
standalone & gpt-oss-20b & 500 & 72.2 & 3.3 & 0.00276 & 0.113 & 100 & 0.659 \\
standalone & llama3.3-70b & 100 & 80.0 & 7.8 & 0.01440 & 0.496 & 100 & 0.440 \\
standalone & qwen3.5-35b-a3b & 500 & 83.4 & 4.3 & 0.00280 & 0.145 & 100 & 0.716 \\
\bottomrule
\end{tabularx}
\end{table}

\section{Architecture-Family Analysis}
\label{sec:arch-analysis}

\subsection{Standalone Baselines}

Standalone results confirm that model family, size, and quantization all contribute to benchmark performance in task-dependent ways. The MoE checkpoints \texttt{gemma4-26b-a4b} and \texttt{qwen3.5-35b-a3b} are particularly competitive: they match or exceed larger dense models on several benchmarks while operating at latencies of 2.8--7.4 seconds and costs of \$0.00185--\$0.00482 per query, yielding EATS of roughly 0.65--0.86 across benchmarks---the best standalone efficiency in the study. \texttt{llama3.3-70b} consistently achieves the lowest EATS among standalone models (0.44--0.53, at \$0.01630--\$0.02034 per query) because its infrastructure cost dominates the EATS penalty term. The dense \texttt{gemma4-31b} is the strongest dense standalone, peaking at 97.0\% on ARC and reaching EATS 0.67--0.72 across benchmarks, whereas \texttt{qwen3.5-27b} attains comparable accuracy (up to 95.0\%) but its higher latency and cost limit its EATS to 0.53--0.60.

\texttt{gpt-oss-120b} performs consistently across the suite---MMLU (87.0\%), ARC (95.4\%), GSM8K (96.2\%), and TruthfulQA (82.4\%)---while \texttt{gpt-oss-20b} is strong on ARC (93.8\%) and GSM8K (94.8\%) but weaker on MMLU (79.8\%) and TruthfulQA (72.2\%). An earlier evaluation snapshot reported sharply lower GSM8K accuracy for both checkpoints; this proved to be an answer-extraction artefact that was corrected by rescoring, as discussed in Section~\ref{sec:exp-discussion}.

\subsection{Routing and Speculative Decoding}

Routing and speculative decoding using the \texttt{gemma4-4b}$\rightarrow$\texttt{gemma4-31b} pair produce the most prominent efficiency findings in this study. Speculative decoding records the highest EATS on four of the five benchmarks---MMLU (0.823), ARC (0.908), HellaSwag (0.899), and TruthfulQA (0.860)---with routing the runner-up on each (0.798, 0.904, 0.893, and 0.854). The ARC speculative result (0.908) is the single highest EATS recorded anywhere in the study. On ARC, HellaSwag, and TruthfulQA these hybrid values exceed the best standalone EATS by 5--14 points, combining near-monolithic accuracy with sub-2-second algorithmic latency and per-query costs below \$0.001. Speculative decoding matches or slightly exceeds routing accuracy on these benchmarks (for example 94.0\% versus 91.0\% on ARC, and 92.0\% versus 90.0\% on HellaSwag). Although speculative decoding invokes the LLM on every query (LLM\,\% = 100), it remains efficient because the verifier rewrites only the divergent suffix of the SLM draft rather than regenerating the full answer from scratch; the resulting reduction in output tokens keeps per-query cost and latency substantially below standalone LLM baselines.

The LLM-invocation rate of routing varies considerably across benchmarks---56\% on MMLU, 75\% on ARC, 81\% on TruthfulQA, 91\% on HellaSwag, and only 1\% on GSM8K---reflecting task-dependent differences in SLM confidence at the fixed threshold of 0.8. Tasks such as GSM8K, where \texttt{gemma4-4b} produces high-confidence arithmetic traces, rarely trigger escalation; commonsense continuation tasks such as HellaSwag, where confidence calibration is weaker, escalate the majority of queries. This variation confirms that the LLM-call ratio is governed by the interaction between task format, SLM capability, and the confidence threshold, rather than being a fixed property of the routing architecture.

On MMLU, speculative decoding reaches the highest EATS (0.823) at 82.0\% accuracy, with routing close behind (0.798 at 77.0\%). These results show that the hybrid pair leads the efficiency frontier even on the broadest world-knowledge workload in the study, though it no longer sets the top raw-accuracy point (the standalone \texttt{gemma4-31b} reaches 87.8\%).

On GSM8K, routing escalates only 1\% of queries to the LLM and achieves 95.0\% accuracy at EATS 0.776. This indicates that \texttt{gemma4-4b} is self-sufficient for nearly all GSM8K questions at the configured confidence threshold, reducing the LLM-call overhead nearly entirely. The higher algorithmic latency (12.1 seconds) relative to ARC/HellaSwag/TruthfulQA reflects longer arithmetic reasoning output lengths rather than orchestration overhead; on this benchmark the lower-latency Pure Swarm overtakes both hybrids on EATS.

\subsection{Active Oracle}

Active Oracle is now evaluated on all five benchmarks, achieving 72.0\% (MMLU), 91.0\% (ARC), 92.0\% (GSM8K), 67.0\% (HellaSwag), and 72.0\% (TruthfulQA) accuracy with EATS of 0.636--0.886. Following the chain-of-thought minimisation described in Section~\ref{sec:taxonomy}, its algorithmic latency fell to 2.8--12.9 seconds, and the architecture is now efficiency-competitive: on ARC it reaches EATS 0.886, the third-highest on that benchmark and above every standalone checkpoint, trailing only routing and speculative decoding. It nonetheless does not overtake the hybrid pair on any benchmark, and its accuracy on commonsense completion (HellaSwag, 67.0\%) and broad knowledge (MMLU, 72.0\%) remains modest. A key structural reason is that the sub-question is generated from the SLM's current, partially incorrect intermediate state; the oracle response therefore sometimes addresses a mis-specified query and fails to redirect the SLM's reasoning chain toward the correct answer, which limits its gains on tasks where the SLM is broadly rather than locally uncertain.

\subsection{Debate (POA)}

Debate is evaluated in two configurations on every benchmark: an all-SLM variant in which a \texttt{gemma4-4b} arbitrator decides between the Proponent and Opponent, and a hybrid variant in which arbitration is delegated to \texttt{gemma4-31b}. The arbitrator choice dominates the outcome. The LLM-arbitrated variant reaches 80.0--96.0\% accuracy and outperforms the all-SLM variant on every benchmark, with the largest gaps on HellaSwag (90.0\% vs.\ 72.0\%) and TruthfulQA (87.0\% vs.\ 72.0\%)---precisely the tasks where small-model judgement is least reliable. On structured reasoning tasks the gap narrows to one to five points (96.0\% vs.\ 95.0\% on GSM8K), indicating that adversarial critique among SLMs is largely self-sufficient when answers are objectively verifiable. Because every debate requires at least three model calls, EATS remains in the 0.52--0.68 range for both variants, below the leaner hybrid designs.

\subsection{Blackboard and Entropy Blackboard}

The two decentralised task-board architectures are evaluated on all five benchmarks using \texttt{gemma4-4b} workers and a \texttt{gemma4-31b} sweeper. Every reported row is a single 100-sample run at the principal operating point, so each design--policy comparison is a fair, setting-matched pair, and the 2$\times$2 matrix is complete for both architectures on all five benchmarks.

Standard Blackboard is strongest on structured reasoning. Under the highest-bid policy it reaches 95.0\% on ARC (EATS 0.808) and 97.0\% on GSM8K---tying the highest raw accuracy recorded in the study, alongside the first-threshold Entropy Blackboard and the standalone \texttt{gemma4-31b} on ARC---while resolving 83--87\% of queries on the SLM workers alone. With the matrix now complete, the claim-policy effect proves modest on every benchmark: highest-bid claiming wins narrowly on accuracy on ARC (95.0\% vs.\ 92.0\%), GSM8K (97.0\% vs.\ 96.0\%), HellaSwag (79.0\% vs.\ 77.0\%), and TruthfulQA (69.0\% vs.\ 68.0\%), while first-threshold claiming wins on MMLU (79.0\% vs.\ 77.0\%); the accuracy gap never exceeds three points. On EATS the two policies are likewise close and benchmark-dependent---first-threshold is the more efficient standard-Blackboard policy on MMLU, GSM8K, and TruthfulQA, and highest-bid on ARC and HellaSwag---with the difference small in every case. TruthfulQA is the principal failure case at this operating point---both policies land at 68.0--69.0\%, well below the routing pair---but the near-tie between policies shows the limitation is the task-board contract on factual calibration, not the claim rule. This corrects an earlier reading based on legacy runs: an exploratory first-threshold run at the old settings (bid threshold 0.65, 1.5-second time-to-live) reached 83.0\% on TruthfulQA by escalating every query to the sweeper, effectively LLM fallback, and a matching legacy run reached 95.0\% on GSM8K; both conflate the claim policy with the legacy threshold and time-to-live and are excluded from the tables, with the matched runs above now providing the controlled comparison.

The bid threshold itself was explored on MMLU before the principal operating point was fixed. Under the highest-bid policy, three otherwise identical 100-sample runs at bid thresholds of 0.65, 0.75, and 0.85 produced accuracies of 74.0\%, 77.0\%, and 79.0\%, with sweeper escalation rising from 20\% to 30\% to 42\%; a set of 20-sample pilot runs at threshold 0.8 (80.0--85.0\% accuracy) was consistent with this trend. Raising the bid threshold makes workers more reluctant to claim tasks, shifting work to the sweeper and trading cost for accuracy---a monotone relationship that motivates the systematic parameter sweep planned in Section~\ref{sec:pending}. The 0.75 setting was adopted as the principal operating point for all subsequent Blackboard and Entropy Blackboard runs; only the 0.75 run from this exploration appears in the MMLU table, the off-point runs being reported here.

Entropy Blackboard isolates the effect of the entropy penalty, since it differs from standard Blackboard only in the bidding rule. The penalty pays off most clearly where worker confidence is poorly calibrated: on HellaSwag it raises sweeper escalation from 41\% to 82\% and lifts accuracy five to seven points (84.0\% vs.\ 77.0--79.0\%), on MMLU it adds three to six points (82.0--83.0\% vs.\ 77.0--79.0\%), and on TruthfulQA it recovers four to seven points over standard Blackboard (73.0--75.0\% vs.\ 68.0--69.0\%), though all blackboard-family variants remain below routing there. On GSM8K, the first-threshold Entropy Blackboard reaches 97.0\% with only 7\% escalation, tying the highest-bid Blackboard for the highest raw accuracy in the study while yielding EATS 0.826---the most efficient blackboard-family result in the evaluation, just ahead of the first-threshold standard Blackboard on the same benchmark (0.818). ARC is the one benchmark where the penalty gives nothing back, costing two points under the highest-bid policy (93.0\% vs.\ 95.0\%) and matching the standard design under first-threshold claiming (92.0\%).

A consistent picture emerges across the complete matrix: at the principal operating point, the claim policy is a minor knob for both architectures---standard Blackboard varies by at most three points between policies and Entropy Blackboard by at most two, and on HellaSwag the two Entropy Blackboard policies produced identical claim and escalation decisions on all 100 queries at zero temperature. The large policy-dependent swings seen earlier were artefacts of the legacy time-to-live, not of the claim rule. What the entropy penalty contributes is therefore accuracy on poorly calibrated tasks rather than policy robustness: by suppressing overconfident bids before the claim decision is resolved, it escalates the right queries on HellaSwag, MMLU, and TruthfulQA, where standard Blackboard's workers are most likely to claim tasks they cannot answer.

\subsection{Ensemble}

The four-SLM ensemble (\texttt{gemma4-4b + llama3.2-3b + ministral3-3b + qwen3.5-4b}) is evaluated on all five benchmarks. Accuracy ranges from 64.0\% (TruthfulQA) to 96.0\% (GSM8K). The GSM8K result trails only the two 97.0\% blackboard-family configurations among coordination-based designs and the strongest standalone checkpoint (\texttt{qwen3.5-35b-a3b}, 96.4\%) by less than half a point, consistent with ensemble-diversity findings in the literature. However, EATS is limited to 0.44--0.62 across benchmarks because four parallel SLM calls generate latency of 15.6--21.6 seconds and combined energy that heavily penalizes the composite score. Single-SLM ensemble variants on MMLU (\texttt{ministral3-3b} at EATS 0.639, \texttt{qwen3.5-4b} at 0.616) show that a lean single-model ensemble can be more efficient than the full multi-model aggregation when the model has low latency and cost, though accuracy is limited (57.0\% and 75.0\% respectively).

\subsection{Pure Swarm}

Pure Swarm with \texttt{gemma4-4b} uses no heavy-model calls at inference time. It achieves 92.0\% on ARC, 94.0\% on GSM8K, 75.0\% on MMLU, 67.0\% on TruthfulQA, and 65.0\% on HellaSwag. With its coordination latency reduced to 7.0--12.3 seconds, Pure Swarm is now efficiency-competitive rather than latency-bound: it records the highest EATS of any configuration on GSM8K (0.841) and reaches EATS 0.818 on ARC, with values spanning 0.604--0.841 across the suite. The strong performance on ARC and GSM8K suggests that compact peer-agent consensus is effective for structured reasoning tasks with clear verifiable answers, matching the accuracy of much larger standalone models at a fraction of their cost. Pure Swarm's lower scores on HellaSwag and TruthfulQA reflect that commonsense completion and factual calibration are harder for compact peer-agent coordination than science-knowledge or arithmetic tasks.

\section{Cross-Benchmark EATS Comparison}
\label{sec:cross-bench}

Table~\ref{tab:eats-leaders} summarizes the top-three configurations by EATS on each benchmark, providing a compact view of the efficiency--quality frontier across the evaluated architectures.

\begin{table}[htbp]
\centering
\caption{Top-three configurations by EATS per benchmark.}
\label{tab:eats-leaders}
\small
\renewcommand{\arraystretch}{1.2}
\begin{tabularx}{\linewidth}{@{} l X r r r @{}}
\toprule
\textbf{Benchmark} & \textbf{Architecture --- Model} & \textbf{Acc\,\%} & \textbf{EATS} & \textbf{Rank} \\
\midrule
MMLU & speculative --- gemma4-4b $\rightarrow$ gemma4-31b & 82.0 & 0.823 & 1 \\
MMLU & routing --- gemma4-4b $\rightarrow$ gemma4-31b & 77.0 & 0.798 & 2 \\
MMLU & standalone --- qwen3.5-35b-a3b & 86.2 & 0.782 & 3 \\
\midrule
ARC & speculative --- gemma4-4b $\rightarrow$ gemma4-31b & 94.0 & 0.908 & 1 \\
ARC & routing --- gemma4-4b $\rightarrow$ gemma4-31b & 91.0 & 0.904 & 2 \\
ARC & active oracle --- gemma4-4b $\rightarrow$ gemma4-31b & 91.0 & 0.886 & 3 \\
\midrule
GSM8K & pure swarm --- gemma4-4b & 94.0 & 0.841 & 1 \\
GSM8K & entropy\_bb --- gemma4-4b $\rightarrow$ gemma4-31b (first thr.) & 97.0 & 0.826 & 2 \\
GSM8K & standalone --- qwen3.5-35b-a3b & 96.4 & 0.824 & 3 \\
\midrule
HellaSwag & speculative --- gemma4-4b $\rightarrow$ gemma4-31b & 92.0 & 0.899 & 1 \\
HellaSwag & routing --- gemma4-4b $\rightarrow$ gemma4-31b & 90.0 & 0.893 & 2 \\
HellaSwag & standalone --- qwen3.5-35b-a3b & 92.6 & 0.807 & 3 \\
\midrule
TruthfulQA & speculative --- gemma4-4b $\rightarrow$ gemma4-31b & 88.0 & 0.860 & 1 \\
TruthfulQA & routing --- gemma4-4b $\rightarrow$ gemma4-31b & 85.0 & 0.854 & 2 \\
TruthfulQA & standalone --- qwen3.5-35b-a3b & 83.4 & 0.716 & 3 \\
\bottomrule
\end{tabularx}
\end{table}

The table reveals a clear cross-benchmark pattern. Speculative decoding and routing occupy the top two EATS positions on four of the five benchmarks (MMLU, ARC, HellaSwag, and TruthfulQA), with speculative ahead of routing in every case; on ARC, Active Oracle joins them in third. GSM8K is the exception: the lower-latency coordination designs take the top two places---Pure Swarm (0.841) and the first-threshold Entropy Blackboard (0.826)---with the most efficient MoE standalone, \texttt{qwen3.5-35b-a3b} (0.824), a close third. There the routing LLM-call ratio is only 1\%, so the SLM draft pass adds latency without meaningful cost reduction, and the long arithmetic reasoning traces penalise the per-query-cost-heavy hybrids more than the SLM-only swarm. On MMLU, speculative decoding and routing again take the top two EATS positions even though the standalone \texttt{gemma4-31b} retains the highest raw accuracy, because the hybrid pair's substantial cost and latency advantages are weighted heavily in the EATS penalty term.

\section{Discussion}
\label{sec:exp-discussion}

\subsection{Hybrid architectures on the efficiency--quality frontier}
The ARC, HellaSwag, and TruthfulQA results provide strong empirical support for the central hypothesis that routing and speculative decoding can achieve accuracy close to or matching the best standalone models while reducing per-query cost and latency by factors of 2--15 relative to the same LLM running in standalone mode, with comparable reductions in the estimated energy and carbon figures (Section~\ref{sec:exp-discussion}). This directly addresses RQ1 (can collaborative architectures match standalone LLM accuracy?) and RQ2 (what are the efficiency trade-offs?). EATS as a composite signal (RQ3) confirms the joint finding: the hybrid pair occupies the top two EATS positions on ARC, HellaSwag, and TruthfulQA (and on MMLU), with values in the 0.854--0.908 range compared to 0.716--0.859 for the best standalone options.

\subsection{GSM8K GPT-family anomaly: resolved}
An earlier evaluation snapshot reported \texttt{gpt-oss-120b} at 67.8\% and \texttt{gpt-oss-20b} at 59.4\% on GSM8K, sharply inconsistent with their performance on every other benchmark. The suspected cause---answer-extraction sensitivity---was confirmed: GSM8K answers require identifying a numeric value from a free-form response, and the GPT-family models format their reasoning and final answers in ways that defeated the platform's original numeric extraction heuristic. The extraction logic and the answer-format prompt instructions were corrected, and all affected GSM8K runs were rescored from their stored raw responses without re-running inference. Under the corrected scoring, \texttt{gpt-oss-120b} reaches 96.2\% and \texttt{gpt-oss-20b} 94.8\%, in line with their results elsewhere in the suite. The episode illustrates a broader methodological point: free-form benchmarks couple model capability to harness-side parsing, so capability claims on such benchmarks require validating the extraction path per model family.

\subsection{Decentralised coordination and uncertainty-aware bidding}
The newly benchmarked Blackboard family provides the first empirical evidence for the two novel decentralised architectures introduced in Chapter~3. Bid-based task allocation with an LLM sweeper is clearly viable: on ARC and GSM8K the family reaches 92.0--97.0\% accuracy while resolving 72--93\% of queries on SLM workers alone, and the 97.0\% on GSM8K---reached by both the highest-bid Blackboard and the first-threshold Entropy Blackboard---ties for the highest raw accuracy recorded in the study (alongside the standalone \texttt{gemma4-31b} on ARC). Because every reported configuration is a single run at the principal operating point and the design--policy matrix is now complete, two fair comparisons emerge. First, the claim policy is a minor knob: standard Blackboard varies by at most three points between first-threshold and highest-bid claiming on every benchmark, and Entropy Blackboard by at most two. The dramatic policy swings reported in an earlier snapshot---most notably an 83.0\% first-threshold result on TruthfulQA against 69.0\% for highest-bid---turned out to be artefacts of the legacy short time-to-live, which forced near-universal sweeper escalation; under matched settings both policies land at 68.0--69.0\% there. Second, the entropy penalty---the only difference between the two architectures---is what actually moves accuracy: it recovers three to seven points on the benchmarks where confidence is least trustworthy (HellaSwag, MMLU, and TruthfulQA) by escalating the queries that overconfident workers would otherwise claim. The preliminary MMLU bid-threshold exploration points the same way: escalation and accuracy respond monotonically to the bid threshold (74.0\% to 79.0\% as the threshold rises from 0.65 to 0.85), confirming that the bidding rule, not the task board itself, is the architecture's principal tuning surface. This is direct evidence for the design hypothesis, drawn from the uncertainty-aware inference literature, that the practical challenge is deciding when a small model should abstain rather than merely measuring its confidence. The strength of LLM-arbitrated Debate on TruthfulQA (87.0\%) relative to every task-board variant suggests that factual calibration benefits from adversarial critique of a candidate answer more than from reallocating the question to a stronger model.

\subsection{Pure Swarm capability ceiling}
Pure Swarm demonstrates that compact SLM peer coordination can reach 92.0\% on ARC and 94.0\% on GSM8K with zero LLM calls, partially addressing RQ4 by matching multi-SLM collaboration findings from the reviewed literature. With its coordination latency now reduced to 7--12 seconds, the design is no longer latency-bound: it records the highest EATS on GSM8K (0.841) and the third-highest on ARC, showing that an SLM-only swarm can lead the efficiency frontier on structured-reasoning workloads. Its weaker accuracy on commonsense completion (HellaSwag) and factual calibration (TruthfulQA) marks the remaining capability ceiling of LLM-free coordination.

\subsection{Ensemble accuracy ceiling}
The four-SLM ensemble achieves 96.0\% on GSM8K, jointly with LLM-arbitrated Debate and one point behind the two blackboard-family configurations at 97.0\%, consistent with ensemble-diversity findings in the literature. EATS is limited to 0.619 there because four parallel SLM calls carry a latency and energy penalty that the EATS formulation weights heavily. This result illustrates the fundamental trade-off the EATS metric is designed to quantify: accuracy gains from coordination are real but arrive at a cost that may not be acceptable in latency-sensitive deployments.

\section{Planned Extensions}
\label{sec:pending}

The current results cover all five benchmarks in the evaluation suite (MMLU, ARC-Challenge, GSM8K, HellaSwag, and TruthfulQA), every architecture defined in the Chapter~3 taxonomy, and the complete blackboard-family design--policy matrix of \{Blackboard, Entropy Blackboard\} $\times$ \{first threshold, highest bid\} at the principal operating point. Two extensions are identified as future work.

First, the sample counts for the collaborative architectures will be deepened: every hybrid, ensemble, swarm, and task-board configuration currently rests on a single 100-sample run, compared with 500 samples for the principal standalone baselines; additional runs will tighten confidence intervals and allow seed-level variance to be reported. Second, the sensitivity of the blackboard family to its orchestration parameters---the bid threshold, the time-to-live before sweeper fallback, and the entropy-penalty weight---will be characterised systematically. The preliminary MMLU bid-threshold exploration (Section~\ref{sec:arch-analysis}) already shows a monotone escalation--accuracy response to the bid threshold, and the time-to-live has proven consequential enough to have confounded an earlier reading of the TruthfulQA results; extending the sweep to all five benchmarks, to the time-to-live, and to the entropy-penalty weight will map the accuracy--efficiency balance these parameters govern.
