\chapter*{Abstract}

Large language models (LLMs) deliver strong capabilities on reasoning and coding tasks but impose substantial inference costs in latency, energy, and monetary expenditure. While recent work explores collaborative architectures that combine small language models (SLMs) with LLMs through routing, verification, debate, and decentralized coordination, the research community still lacks a systematic evaluation framework that treats cost and energy as first-class outcomes alongside accuracy. This gap was identified through a companion systematic literature review of 54 collaborative SLM--LLM studies, which directly motivated the measurement-first design of the present platform.

This report presents an experiment platform in which every architecture consumes a normalized \texttt{Query} and returns a normalized \texttt{Response}, enabling direct comparison across answer accuracy, latency, token usage, cost, energy consumption, carbon emissions, LLM-call ratio, and the Efficiency-Accuracy Trade-off Score (EATS). The platform spans three GPU tiers---dedicated Google Cloud L4 instances for SLMs, a shared RTX PRO6000 host for mid-tier models, and a Nebius H200 host for frontier-scale inference---and integrates a proxy-based energy and carbon instrumentation layer with MLflow-backed experiment tracking. The architecture taxonomy covers three families: monolithic baselines (including Mixture-of-Experts), hybrid collaboration strategies (routing, speculative decoding, debate, active oracle, blackboard, and entropy blackboard), and multi-SLM coordination designs (ensemble and pure swarm).

As preliminary validation of the platform's measurement capability, two pilot comparisons illustrate operating-point trade-offs within the broader nine-architecture evaluation space. On MMLU, a routing configuration with \texttt{gemma4-4b} reached 76\% accuracy against an 85\% monolithic \texttt{gemma4-31b} baseline with only 34\% heavy-model usage, reducing energy and CO$_2$ by approximately 20\% at the cost of 2.2$\times$ higher average latency. On GSM8K, the same routing strategy reached 91\% accuracy---between the SLM-only baseline at 83\% and the LLM baseline at 95\%---while reducing cost by 25\% and energy by 42\%. These two data points represent only the routing-versus-monolithic slice of the study; the remaining architectures (speculative decoding, debate, active oracle, blackboard, entropy blackboard, ensemble, and pure swarm), the full five-benchmark suite covering broad knowledge, arithmetic, commonsense, and truthfulness, and twelve model checkpoints across three GPU tiers constitute the ongoing comparative evaluation.

These results confirm that no single architecture dominates across all metrics simultaneously and that the platform successfully captures meaningful quality-efficiency trade-offs. The current evidence remains exploratory: the intended coding benchmark, the swarm architecture family, and the complete end-to-end validation pipeline are still in progress. The platform's primary contribution at this stage is a reusable, measurement-complete infrastructure for the comparative evaluation phase of the thesis.

\vspace{0.5cm}

\noindent \textbf{Keywords:} small language models, large language models, orchestration architectures, energy-aware evaluation, efficiency-accuracy trade-off
