\appendix
\chapter{APPENDIX}

\section{Canonical Model Aliases}

Table~\ref{tab:appendix-aliases} lists the canonical aliases used by the control plane. These aliases are important because they decouple experiment configuration from the served model identifiers exposed by remote vLLM endpoints. All models are open-weight checkpoints available on Hugging Face. ${}^{a}$\texttt{phi4-mini} is part of the registered model pool but was not used in any experiment reported in this term report.

\begin{longtable}{@{} l l l @{}}
    \caption{Canonical model aliases used by the platform.}
    \label{tab:appendix-aliases} \\
    \toprule
    \textbf{Alias} & \textbf{Tier} & \textbf{Typical host class} \\
    \midrule
    \endfirsthead
    \toprule
    \textbf{Alias} & \textbf{Tier} & \textbf{Typical host class} \\
    \midrule
    \endhead
    \texttt{gemma4-4b} & SLM & Dedicated GCP L4 \\
    \texttt{qwen3.5-4b} & SLM & Dedicated GCP L4 \\
    \texttt{llama3.2-3b} & SLM & Dedicated GCP L4 \\
    \texttt{ministral3-3b} & SLM & Dedicated GCP L4 \\
    \texttt{phi4-mini}${}^{a}$ & SLM & Dedicated GCP L4 \\
    \texttt{gpt-oss-20b} & Mid-tier LLM & Shared RTX PRO6000 \\
    \texttt{qwen3.5-27b} & Mid-tier LLM & Shared RTX PRO6000 \\
    \texttt{gemma4-31b} & Mid-tier LLM & Shared RTX PRO6000 \\
    \texttt{qwen3.5-35b-a3b} & Mid-tier MoE & Shared RTX PRO6000 \\
    \texttt{gemma4-26b-a4b} & Mid-tier MoE & Shared RTX PRO6000 \\
    \texttt{llama3.3-70b} & Heavy LLM & Nebius heavy host \\
    \texttt{gpt-oss-120b} & Heavy LLM & Nebius heavy host \\
    \bottomrule
\end{longtable}

\section{Host Topology Summary}

The current host topology used by the platform can be summarized as follows:

\begin{itemize}
    \item one dedicated Google Cloud L4 host per SLM alias,
    \item one shared Google Cloud RTX PRO6000 host for the 20B--40B class models,
    \item one optional shared Nebius heavy host for the 70B and 120B class models, and
    \item a local control plane that communicates with all remote hosts through OpenAI-compatible HTTP endpoints.
\end{itemize}

\section{Human Preference Benchmark Record Schema}

The intended record shape for the human preference benchmark, planned as a complementary open-ended evaluation surface (see Section~5.4), is:

\begin{itemize}
    \item \texttt{prompt\_id} and \texttt{prompt\_text}
    \item \texttt{architecture\_a}, \texttt{architecture\_b}
    \item \texttt{answer\_a}, \texttt{answer\_b}
    \item randomized display order
    \item selected winner or tie/skip
    \item evaluator/session identifier and timestamp
    \item latency, token, and cost metadata for both compared answers
\end{itemize}

\section{EATS Definition}

The project-specific EATS metric is defined as:

\[
\text{EATS} =
\frac{\text{accuracy}}
{\text{accuracy} +
 0.40\left(
 0.65\,\text{normalized\_cost} +
 0.20\,\text{normalized\_algorithmic\_latency} +
 0.15\,\text{normalized\_energy}
 \right)
 + 0.60\left(1 - \text{accuracy}\right)}
\]

Its purpose is to reward architectures that preserve useful answer quality while reducing normalized cost, algorithmic latency, and energy relative to a full-LLM baseline. The weighted additive efficiency penalty (weights summing to 1: cost 65\%, latency 20\%, energy 15\%) makes total cost the dominant efficiency dimension, while the explicit accuracy-deficit term $1-\text{accuracy}$ prevents low-accuracy systems from scoring near 1.0 solely because they are cheap. The present revision uses an efficiency scale of $\lambda = 0.40$ and an accuracy-deficit scale of $\beta = 0.60$. Two normalisation modes (per-experiment and aggregate) are defined in Section~3.4.5 of the main text; the result tables in Chapter~4 use aggregate mode.

\section{Example Experiment Configuration Placeholder}

Listing \ref{lst:config-placeholder} shows a minimal experiment configuration placeholder illustrating how an automated run can be described.

\begin{lstlisting}[caption={Example experiment configuration placeholder.}, label={lst:config-placeholder}, language={}]
architecture: routing
benchmark: mmlu
n_samples: 100
slm: qwen3.5-4b
llm: gpt-oss-20b
confidence_threshold: 0.7
output_dir: results
mlflow_tracking_uri: http://localhost:5000
\end{lstlisting}

The final appendix can later be extended with completed experiment configurations, run identifiers, result tables, and final figure references once the benchmark and experiment pipeline are stable.
