\begin{thebibliography}{99}
\addcontentsline{toc}{chapter}{REFERENCES}

\bibitem{slrmanuscript2026}
Project Team, \textit{Collaborative SLM--LLM Systems: A Systematic Literature Review}, manuscript submitted to Euromicro, 2026.

\bibitem{page2021prisma}
Page, M.J., McKenzie, J.E., Bossuyt, P.M., Boutron, I., Hoffmann, T.C., Mulrow, C.D., et al., ``The PRISMA 2020 Statement: An Updated Guideline for Systematic Reviews,'' \textit{BMJ}, vol. 372, n71, 2021.

\bibitem{energytoken2025}
Wilhelm, P., Wittkopp, T., and Kao, O., ``Beyond Test-Time Compute Strategies: Advocating Energy-per-Token in LLM Inference,'' in \textit{Proceedings of the 5th Workshop on Machine Learning and Systems (EuroMLSys 2025)}, pp. 208--215, 2025.

\bibitem{kwon2023vllm}
Kwon, W., Li, Z., Zhuang, S., Sheng, Y., Zheng, L., Yu, C.H., Gonzalez, J.E., Zhang, H., and Stoica, I., ``Efficient Memory Management for Large Language Model Serving with PagedAttention,'' arXiv preprint arXiv:2309.06180, 2023.

\bibitem{hao2024hybrid}
Hao, Z., Wei, J., Wang, T., and Ren, J., ``Hybrid SLM and LLM for Edge-Cloud Collaborative Inference,'' in \textit{Proceedings of the 2024 Workshop on Edge and Mobile Foundation Models (EdgeFM 2024)}, pp. 36--41, 2024.

\bibitem{shao2025dot}
Shao, C., Hu, X., Lin, Y., and Xu, F., ``Division-of-Thoughts: Harnessing Hybrid Language Model Synergy for Efficient On-Device Agents,'' in \textit{Proceedings of the ACM Web Conference 2025}, pp. 1822--1833, 2025.

\bibitem{multiagent2025}
Zeng, J., Liao, J., Sun, S., Zhang, G., Xu, C., Liu, H., et al., ``Multi-Agent Collaborative Reasoning via Cloud-Edge Framework,'' \textit{IEEE Network}, 2026.

\bibitem{lee2024orchestrallm}
Lee, C.H., Cheng, H., and Ostendorf, M., ``OrchestraLLM: Efficient Orchestration of Language Models for Dialogue State Tracking,'' in \textit{Proceedings of NAACL-HLT 2024}, pp. 1434--1445, 2024.

\bibitem{oh2025uncertainty}
Oh, S., Kim, J., Park, J., Ko, S.W., Quek, T.Q.S., and Kim, S.L., ``Uncertainty-Aware Hybrid Inference with On-Device Small and Remote Large Language Models,'' in \textit{Proceedings of the 2025 IEEE International Conference on Machine Learning for Communication and Networking}, pp. 1--7, 2025.

\bibitem{citer2025}
She, J., Zheng, W., Liu, Z., Wang, H., Xing, E.P., Yao, H., et al., ``Token Level Routing Inference System for Edge Devices,'' in \textit{Proceedings of the 63rd Annual Meeting of the Association for Computational Linguistics: System Demonstrations}, pp. 159--166, 2025.

\bibitem{rethinkllm2025}
Alabbasi, N., Alhussein, O., and Muhaidat, S., ``Do You Actually Need an LLM? Rethinking Language Models for Customer Reviews Analysis,'' arXiv preprint arXiv:2411.02555, 2024.

\bibitem{alabbasi2024teleoracle}
Alabbasi, N., Erak, O., Alhussein, O., Lotfi, I., Muhaidat, S., and Debbah, M., ``TeleOracle: Fine-Tuned Retrieval-Augmented Generation with Long-Context Support for Networks,'' arXiv preprint arXiv:2411.02617, 2024.

\bibitem{debateagents2025}
Huynh, T., and Nguyen, D., ``Collaborating Agents with Large Language Models for Question-Answering Tasks,'' in \textit{Proceedings of the 5th IEEE International Conference on Software Engineering and Artificial Intelligence (SEAI)}, pp. 1--5, 2025.

\bibitem{resiliomate2025}
Xiong, Y., Huang, M., Liang, X., and Tao, M., ``ResilioMate: A Resilient Multi-Agent Task Executing Framework for Enhancing Small Language Models,'' \textit{IEEE Access}, vol. 13, pp. 86892--86911, 2025.

\bibitem{bioagents2025}
Mehandru, N., Hall, A.K., Melnichenko, O., Dubinina, Y., Tsirulnikov, D., Bamman, D., et al., ``BioAgents: Bridging the Gap in Bioinformatics Analysis with Multi-Agent Systems,'' \textit{Scientific Reports}, vol. 15, 39036, 2025.

\bibitem{sentensemble2026}
Cielen, D., De Bock, K.W., and Florès, L., ``Exploring Zero-Shot SLM Ensembles as an Alternative to LLMs for Sentiment Analysis,'' \textit{Information Fusion}, vol. 126, 103666, 2026.

\bibitem{compos2024}
Das, S., and Srihari, R., ``Compos Mentis at SemEval-2024 Task 6: A Multi-Faceted Role-Based Large Language Model Ensemble to Detect Hallucination,'' in \textit{Proceedings of the 18th International Workshop on Semantic Evaluation (SemEval-2024)}, pp. 1449--1454, 2024.

\bibitem{hendrycks2021mmlu}
Hendrycks, D., Burns, C., Basart, S., Zou, A., Mazeika, M., Song, D., and Steinhardt, J., ``Measuring Massive Multitask Language Understanding,'' in \textit{Proceedings of the International Conference on Learning Representations (ICLR)}, 2021.

\bibitem{clark2018arc}
Clark, P., Cowhey, I., Etzioni, O., Khot, T., Sabharwal, A., Schoenick, C., and Tafjord, O., ``Think You Have Solved Question Answering? Try ARC, the AI2 Reasoning Challenge,'' arXiv preprint arXiv:1803.05457, 2018.

\bibitem{zellers2019hellaswag}
Zellers, R., Holtzman, A., Bisk, Y., Farhadi, A., and Choi, Y., ``HellaSwag: Can a Machine Really Finish Your Sentence?,'' in \textit{Proceedings of the 57th Annual Meeting of the Association for Computational Linguistics}, pp. 4791--4800, 2019.

\bibitem{cobbe2021gsm8k}
Cobbe, K., Kosaraju, V., Bavarian, M., Chen, M., Jun, H., Kaiser, L., Plappert, M., Tworek, J., Hilton, J., Nakano, R., Hesse, C., and Schulman, J., ``Training Verifiers to Solve Math Word Problems,'' arXiv preprint arXiv:2110.14168, 2021.

\bibitem{lin2022truthfulqa}
Lin, S., Hilton, J., and Evans, O., ``TruthfulQA: Measuring How Models Mimic Human Falsehoods,'' in \textit{Proceedings of the 60th Annual Meeting of the Association for Computational Linguistics}, pp. 3214--3252, 2022.

\end{thebibliography}
