\chapter*{ALIGNMENT WITH UN SUSTAINABLE DEVELOPMENT GOALS}
\addcontentsline{toc}{chapter}{ALIGNMENT WITH UN SUSTAINABLE DEVELOPMENT GOALS}

This project does not target a single social sector such as healthcare or education.
Its contribution is infrastructural: it improves how language-model systems are evaluated,
compared, and governed under a shared measurement policy.
That still creates meaningful alignment with the United Nations Sustainable Development Goals
because the platform makes hidden resource trade-offs visible, discourages wasteful deployment
choices, and lowers the technical and economic barriers to competitive AI development.
The Efficiency-Adjusted Task Score (EATS) makes this alignment concrete by treating
energy and cost as primary ranking criteria rather than secondary diagnostics,
so every architecture comparison the platform produces carries an explicit sustainability signal.

\begin{table}[htbp]
    \centering
    \caption{Alignment of the thesis project with selected UN SDGs.}
    \small
    \begin{tabularx}{\linewidth}{@{} l l X @{}}
        \toprule
        \textbf{SDG} & \textbf{Impact} & \textbf{Project relevance} \\
        \midrule
        SDG 8 --- Decent Work and Economic Growth
            & Direct
            & Cost-aware orchestration and shared-host visibility help engineering teams reason
              about whether a workload truly requires a frontier-scale model.
              Quantifying the cost--quality trade-off improves allocation decisions and reduces
              unnecessary API and infrastructure spending. \\
        \addlinespace
        SDG 9 --- Industry, Innovation \& Infrastructure
            & Direct
            & The project delivers an open, reproducible experiment platform for AI system evaluation.
              Common \texttt{Query}/\texttt{Response} contracts, shared benchmark tasks, and a
              public reporting schema lower the cost of replication and extend the evaluation
              infrastructure available to the research community. \\
        \addlinespace
        SDGs 12 \& 13 --- Responsible Consumption \& Climate Action
            & Direct
            & Energy and carbon tracking (\texttt{energy\_kwh}, \texttt{co2\_g}) are first-class
              platform outputs.
              EATS incorporates both cost and energy into its composite score, so architectures
              are ranked on quality \emph{per unit of environmental cost}, not on accuracy alone.
              This directly incentivises the selection of lower-energy deployment paths. \\
        \addlinespace
        SDG 10 --- Reduced Inequalities
            & Indirect
            & By empirically validating that hybrid and multi-SLM architectures can approach
              monolithic LLM accuracy on reasoning tasks, the platform provides evidence that
              competitive AI development is achievable without frontier-scale hardware or cloud
              budgets, lowering the economic barrier to adoption for resource-constrained teams. \\
        \addlinespace
        SDG 16 --- Peace, Justice and Strong Institutions
            & Indirect
            & Transparent reporting of model behavior, host utilization, token budgets, and
              benchmark assumptions supports accountable evaluation practices.
              Reproducible, open reporting reduces the risk of misleading benchmark claims and
              contributes to norms of integrity in AI system assessment. \\
        \bottomrule
    \end{tabularx}
\end{table}

\section*{SDG 8 and SDG 9: Economic Efficiency and Open Infrastructure}

SDG~8 is addressed through the cost dimension of EATS and through the explicit cost reporting
that accompanies every result.
By quantifying token usage, API cost, and host resource consumption for each architecture,
the platform enables teams to identify when a cheaper, smaller-model solution is sufficient for
a given workload.
The experimental results confirm that hybrid and ensemble architectures can match or closely
approach monolithic LLM accuracy at meaningfully lower cost, providing evidence-based guidance
for engineering investment decisions.

The platform itself is simultaneously a contribution to shared AI evaluation infrastructure,
directly supporting SDG~9.
The common \texttt{Query}/\texttt{Response} contract, the benchmark task library, and the
reporting schema are all designed to be architecture-agnostic and reusable, so that future
studies can extend the platform rather than rebuild it.
This reduces the duplication of effort that currently characterises LLM evaluation work and
lowers the barrier to reproducible AI research for teams without dedicated infrastructure.

\section*{SDGs 12 and 13: Responsible Consumption and Climate Action}

The strongest direct alignment arises from the platform's environmental measurement layer.
Every experiment records \texttt{energy\_kwh} and \texttt{co2\_g} alongside the conventional
accuracy and latency outcomes.
These fields are not optional annotations; they are required fields in the shared
\texttt{Response} schema, so no architecture can be evaluated without also producing an
energy and carbon footprint.
EATS formalises this alignment: because EATS is a composite of accuracy, cost, latency,
and energy, an architecture that consumes excessive energy is penalised in the primary
ranking signal even if its raw accuracy is competitive.
The practical consequence is that every architecture comparison the platform produces is
simultaneously a comparison of environmental impact, making SDG~12 and SDG~13 goals
operational rather than aspirational within the evaluation workflow.
It is worth acknowledging that running benchmarks itself incurs an energy cost; the platform
is designed to execute on shared, multi-tenant infrastructure so that this overhead is
amortized across multiple concurrent experiments rather than dedicated to a single run.

\section*{SDG 10 and SDG 16: Reducing Inequalities and Accountable AI}

The platform's most significant indirect contribution to reduced inequality is the empirical
grounding it provides for SLM-based deployment strategies.
A recurring finding in the systematic literature review (Chapter~2) is that small-model
solutions are rarely benchmarked against frontier models under identical conditions, which
makes it difficult for resource-constrained teams to justify adopting them.
By placing monolithic LLM baselines and hybrid or multi-SLM architectures on the same
evaluation surface, the platform produces direct, comparable evidence for the accuracy--cost
trade-off, reducing the informational disadvantage that teams without frontier-model budgets
currently face.

Alignment with SDG~16 follows from the platform's transparency design.
Benchmark tasks, evaluation prompts, scoring policies, and measurement methodology are all
documented and version-controlled.
Host assignment logs and token budget records make it possible to audit whether any
architecture received a systematic advantage.
These properties support the kind of accountable, reproducible AI evaluation that is a
prerequisite for trustworthy institutional decision-making about which systems to deploy.
